# (PART)  Assessment {-} 

# Learning the Tidyverse2 {#tidyverse2 .unnumbered}

## Visualising Module 1 Flow Data {-}

### Setting up {-}

#### Creating an R Studio project {-}

1) Open R Studio
2) Go *File -> New Project -> New Directory -> New Project*
3) *Directory name:* ENVT3362_workshop_2
4) *Create project as a subdirectory of:* Wherever you store your ENVT3362 files!
5) Click *Create project*
5) Download the spreadsheet for this workshop [here](envFlowData.xls)
6) Move this to your *ENVT3362_workshop_2* directory

### Importing and formatting the data {-}

#### Load the necessary packages {-}

- readxl and lubridate need to be loaded separately 

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(readxl)
library(lubridate)
```

#### Import the spreadsheet {-}

- The `path` argument is relative to you R Studio project file
- `sheet` specifies which Excel sheet to read

```{r message=FALSE, warning=FALSE, include=FALSE}
envFlow  <- read_xls(path = "data/workshop2/envFlowData.xls", sheet = 1)
```

```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
envFlow  <- read_xls(path = "envFlowData.xls", sheet = 1)
```

#### Inspect the data {-}

- `head()` prints the first few observations
- What data type is `date`?


```{r echo=TRUE, message=FALSE, warning=FALSE}
head(envFlow)
```

#### Format the date {-}

- Use lubridate's `ymd()` function to overwrite the existing `date` variable and convert the **character** data to **date** data

```{r echo=TRUE, message=FALSE, warning=FALSE}
envFlow$date <- ymd(envFlow$date)
```

#### Inspect the data again {-}

- Notice the change in data type of `date`

```{r echo=TRUE, message=FALSE, warning=FALSE}
head(envFlow)
```

### Graphing with base plot {-}

Simply using the 'plot' function will make a default points plot. The plot does not look good, however, the function is convenient for checking your data quickly.


#### Call `plot()` {-}

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow) 
```

'plot' defaults to a points plot, but we can quickly change it to a line plot using the argument `type = "l"`.


```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow, type = "l") 
```

Just as with ggplot, we can access the R database of standard colours. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow, type = "l", col = "dodgerblue3") 
```

By default it is selecting x and y from columns 1 and 2. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(x=envFlow$date, y=envFlow$totalDischarge    ,type="l",col="dodgerblue3")
```

If you had several columns, you could choose which to plot on the y axis. We can make a new variable, say, double flow. A new line can be appended to the original plot using the `lines` function.


```{r echo=TRUE, message=FALSE, warning=FALSE}
doubleflow = envFlow$totalDischarge * 2
plot(x=envFlow$date, y=envFlow$totalDischarge    ,type="l",col="dodgerblue3")
lines(x=envFlow$date, y=doubleflow, col="red4")
```

To plot a subset of the data, use the square brackets to select indices, say, points 100 to 800.

```{r echo=TRUE, message=FALSE, warning=FALSE}
doubleflow = envFlow$totalDischarge * 2
plot(x=envFlow$date[100:800], y=envFlow$totalDischarge[100:800]    ,type="l",col="dodgerblue3")
```

Similarly, the limits of the axes can be set with the `xlim` and `ylim` arguments. 

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(x=envFlow$date, y=envFlow$totalDischarge    
     ,type="l",col="dodgerblue3"
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     ,ylim = c(0,1500)
     )
```

#### Call `plot()` and then start again {-}

The base plot is convenient, however, to make it look good, sometimes it is easier to plot it as a blank and add each element in separately. Just as "l" was the argument for lines, "n" is an argument for no plot. We can even turn off the axes and titles.

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow,type="n"
     ,axes=F
     ,xlab="", ylab=""
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     )
```

Then we can add all the elements back in. The plot is still open so it will plot everything to the blank plot.


```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow,type="n"
     ,axes=F
     ,xlab="", ylab=""
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     )
lines(envFlow$date,envFlow$totalDischarge,col="dodgerblue3")
lines(envFlow$date,doubleflow,col="red4")
box(bty="o")
```

#### Axis and mtext {-}
We can put the axis and axis labels back in using the functions 'axis' and 'mtext'. They both have the `side` argument:

1 = bottom, 2 = left, 3 = top, 4 = right

The font size defaults to 1, but we can adjust it using the argument `cex`. The distance from the axis is set using `line`. the `expression` function allows formatting of text, such as subscripts and superscripts.

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow,type="n"
     ,axes=F
     ,xlab="", ylab=""
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     )
lines(envFlow$date,envFlow$totalDischarge,col="dodgerblue3")
lines(envFlow$date,doubleflow,col="red4")
box(bty="o")
mtext(side = 1,text="Date",line=2)
mtext(side = 2,text=expression("Flow (ML d"^-1*")"),line=2)
axis(2,cex=0.7)
```

The date axis needs to be formatted, then the `axis.Date` function is used. This step is clunky compared to ggplot.
```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow,type="n"
     ,axes=F
     ,xlab="", ylab=""
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     )
lines(envFlow$date,envFlow$totalDischarge,col="dodgerblue3")
lines(envFlow$date,doubleflow,col="red4")
box(bty="o")
mtext(side = 1,text="Date",line=2)
mtext(side = 2,text=expression("Flow (ML d"^-1*")"),line=2)
axis(2,cex=0.7)
x.axis<-as.Date(seq(min(envFlow$date),max(envFlow$date),by=3*365),format="%Y")
axis.Date(at=x.axis,side=1,cex=0.7)
```

#### Legend {-} 

Add a legend using the `legend` function. You can search for its details. In this case, `topleft` is the position, inset by 0.1; the `legend` argument is the text, `col` is the sequence of colours, and pch is the symbol.

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(envFlow,type="n"
     ,axes=F
     ,xlab="", ylab=""
     ,xlim = c(date("2000-01-01"),date("2005-01-01"))
     )
lines(envFlow$date,envFlow$totalDischarge,col="dodgerblue3")
lines(envFlow$date,doubleflow,col="red4")
box(bty="o")
mtext(side = 1,text="Date",line=2)
mtext(side = 2,text=expression("Flow (ML d"^-1*")"),line=2)
axis(2,cex=0.7)
x.axis<-as.Date(seq(min(envFlow$date),max(envFlow$date),by=3*365),format="%Y")
axis.Date(at=x.axis,side=1,cex=0.7)
legend("topleft", inset=0.1
       ,legend=c("Flow","Doubleflow")
       ,col=c("dodgerblue3","red4")
       ,pch=16 
       ,cex = 0.7)
```



