# Modelling Quokkas in R {#quokkas -}

### Setting up {-}

#### Create a new R script {-}

- Go *File -> New File -> R Script*
- Load the tidyverse

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

#### Declare the constants as variables {-}

```{r echo=TRUE, message=FALSE, warning=FALSE}
birthRate  <- 1
deathRate  <- 0.1
timestep  <- 1 # i.e. 1 year
timeMax <- 30
carryCap  <- 10000
```

#### Create a new data frame of the initial values {-}

- A data frame is simply a table
- `year`, `relativeCarryCap`, `quokkaRate`, and `quokkaRate` will be created as new columns in that table
- Inspect `model` with the `head()` function

```{r echo=TRUE, message=FALSE, warning=FALSE}
model  <- data.frame(
    year = 1,
    relativeCarryCap  = NA,
    quokkaRate = NA,
    quokkaPop = 5
)

head(model)
```

### Modelling {-}

#### Calculate the next timestep {-}

For the next time step we'll need to compute new values for `year`, `relativeCarryCap`, `quokkaRate`, and `quokkaPop`. These values will be appended as a new row (i.e. timestep) to the `model` data frame.

- `year`takes a value of 2

```{r echo=TRUE, message=FALSE, warning=FALSE}
year <- 2
```

- `relativeCarryCap` is the result of the previous quokka population divided by the carrying capacity
- `model$quokkaPop[1]` specifies the population in the previous row (row 1) i.e. the previous time step

```{r echo=TRUE, message=FALSE, warning=FALSE}
relativeCarryCap  <-  model$quokkaPop[1]/carryCap
```

- `quokkaRate` and `quokkaPop` are calculated in the same fashion i.e. using the constants and values in the previous row of the `model` data frame

```{r echo=TRUE, message=FALSE, warning=FALSE}
quokkaRate <- timestep*((1-relativeCarryCap)*birthRate*model$quokkaPop[1]-(deathRate*model$quokkaPop[1]))

quokkaPop <-  quokkaRate+model$quokkaPop[1]
```

- Next, we'll create an identical data frame to `model` with our new values

```{r echo=TRUE, message=FALSE, warning=FALSE}
model_temp  <- data.frame(
        year = year,
        relativeCarryCap = relativeCarryCap,
        quokkaRate = quokkaRate,
        quokkaPop = quokkaPop
    )

head(model_temp)
```

- We can append this time step to `model` by using the `rbind()` function

```{r echo=TRUE, message=FALSE, warning=FALSE}
model  <- rbind(model,model_temp)
head(model)
```

### Speeding things up {-}

Want to repeat that process for the next 29 time steps? Probably not. Let's create a loop that iterates through each year, calculates the new values, and appends them to the `model` data frame.

```{r echo=FALSE, message=FALSE, warning=FALSE}
model  <- data.frame(
    year = 1,
    relativeCarryCap  = NA,
    quokkaRate = NA,
    quokkaPop = 5
)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
for(i in 2:timeMax){

    year  <- i
    relativeCarryCap  <-  model$quokkaPop[i-1]/carryCap
    quokkaRate = timestep*((1-relativeCarryCap)*birthRate*model$quokkaPop[i-1]-(deathRate*model$quokkaPop[i-1]))
    quokkaPop = quokkaRate+model$quokkaPop[i-1]

    model_temp  <- data.frame(
        year = year,
        relativeCarryCap = relativeCarryCap,
        quokkaRate = quokkaRate,
        quokkaPop = quokkaPop
    )

    model  <- rbind(model,model_temp)

}
```

### Visualise the modelled populations with ggplot {-}

```{r echo=TRUE, message=FALSE, warning=FALSE}
ggplot(data=model,mapping=aes(x=year,y=quokkaPop))+
    geom_line()+
    labs(x="Year", y="Quokka population")+
    theme_light()
```