# Module 5b: Land Use and Nutrient Management {#M5B_1 -} 

### Objectives {#M5B_2 -}

To setup a nutrient export model to compare the effect of Best Management Practices (BMP's) on nutrient loads from a catchment with mixed land use.

### Model Overview {#M5B_3 -}

Consider a catchment basin with with **different types of land use**.

The daily streamflow through the catchment, $Q^t$ (m^3^/day), can be approximated based on the daily rainfall ($R$) - see the Water Balance model in Module 5a. 

The phosphorus concentration in the water, $P^{t}$ (g/m^3), depends on how much flow there is (i.e. more flow = more fertilizer leaching). To model this we use a simple power law to make a "flow-concentration relationship", which is computed for each land-use ($P$ is proportional to $\alpha Q^{\beta}$). *If you've not seen a power law, sketch a graph of C (y-axis) vs Q (x-axis) for a value of $\beta =1$. What happens if $\beta =2$ or $\beta =0.5$?*

This is averaged over the different land use fractions (denoted with the small $p$ index), using the land-use fraction $F_p$:

<center>
\begin{equation}
P^{t} = \sum_{p=4}^{N_{p}} F_{p}\left (\color{#FF0000}{\alpha_{p} \left (\frac{Q^{t}}{A_{c}}  \right )^{\beta_{p}}+P_{0}}  \right )
(\#eq:module5b-3)
\end{equation}
</center>

where alpha ($\alpha$) and beta ($\beta$) are variables that govern the stream $P$ concentration (g/m^3^) as a function of area-averaged flow rate ($Q/A$) and they depend on the dominant land use ($p$). $N_p$ is the number of land use classes being considered (4) and $P_0$ is the background concentration (g/m^3^) for that sub-catchment (i.e. the value of PO~4~ when flow is low 0).


```{block2, hintM5B_1, type='rmdnote2'}
$\color{#FF0000}{\alpha_{p} \left (\frac{Q}{A_{c}}  \right )^{\beta_{p}}+P_{0}}$ is the flow dependent concentration of the *p*-th land use type.
```


The total $P$ export **load** is $P_{load}$ (g/day) and is the mass flux through the river (remember *load = flow x concentration*) :

<center>
\begin{equation}
P_{load}^{t} = P^{t} Q^{t}
(\#eq:module5b-4)
\end{equation}
</center>

<!--
<center>
<video width="80%" height="80%" controls>
  <source src="images/module5b/video1.mp4" type="video/mp4">
</video>
</center>
-->

### Module Resources {#M5B_4 -}

The Excel spreadsheet for this module is part of the Module 5 download. Access using the download button in the tool bar <i class="fa fa-download" aria-hidden="true"></i>.

### Exercises {#M5B_5 -}

#### Building the model {#M5B_6 -}

1) Draw by hand the conceptual model of this system.
2) Confirm with the person beside you what index means ($t$ and $p$).
3) Use your Module 5a water balance model to predict the catchment flows. Tailor that model to use the information as outlined in Table \@ref(tab:module5b-1) below and supplied rainfall data in the **CLIMATE** sheet. Input a value for each parameter within the range given for now.

<center>
```{r module5b-1, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(hablar)
library(dplyr)
#options(knitr.kable.NA = '')
module5b_1 <-
  read.csv("tables/module5b/table1.csv", check.names = FALSE)
kable(
  module5b_1,
  format = "html",
  escape = F,
  align = "c",
  caption = "Parameters for the individual catchments."
) %>%
  kable_styling(full_width = F, position = "center") %>%
  column_spec(1, width_min = "23.5em") %>%
  column_spec(2, width_min = "23.5em") %>%
  #column_spec(1, width_min = "5em") %>%
  #column_spec(2, width_min = "15em") %>%
  #column_spec(3, width_min = "13em") %>%
  #row_spec(1:12, background = '#ffffff00') %>%
  scroll_box(width = "100%",
             height = "100%",
             fixed_thead = FALSE) 
```
</center>


#### Calibrating the model {#M5B_7 -}

Let’s make sure the model flows are realistic. Let's compare our prediction with some observed data:

4) Manually adjust the catchment model parameters (see possible ranges in table below) until the model best matches the observed data - the observed flow data and water table data is in the **CLIMATE** sheet. This processed is called “calibration”.

```{block2, hintM5B_2, type='rmdtip2'}
Check this by using a regression plot of the given total observed streamflow and the model total streamflow, checking to get the highest R^2^ value.
```

```{block2, hintM5B_50, type='rmdtip2'}
Note the catchment area of the WATER BALANCE model needs to match the details in the table. Make a copy of the sheet to make a duplicate model with tailored parameters
```

#### Calculate phosphate concentrations {#M5B_8 -}

Now let’s work out the land-use specific stream $P$ concentrations:

5) Predict the catchment river $P$ concentrations and total export load, $P_{load}$.
The best way to do this is first predict the $P$ concentration of each land-use class each as a column (i.e. Dairy $P$ concentration, Wheat $P$ concentration and so on), and then sum the 4 columns for $P^t$ by factoring the ($F_p$) for each land-use.

<center>
```{r module5b-2, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(hablar)
library(dplyr)
#options(knitr.kable.NA = '')
module5b_2 <-
  read.csv("tables/module5b/table2.csv", check.names = FALSE)
kable(
  module5b_2,
  format = "html",
  escape = F,
  align = "c",
  caption = "Phosphate parameters."
) %>%
  kable_styling(full_width = F, position = "center") %>%
  column_spec(1, width_min = "25em") %>%
  column_spec(2, width_min = "10em") %>%
  column_spec(3, width_min = "10em") %>%
  column_spec(4, width_min = "10em") %>%
  #row_spec(1:12, background = '#ffffff00') %>%
  scroll_box(width = "100%",
             height = "235px",
             fixed_thead = FALSE) 
```
</center>

```{block2, hintM5B_3, type='rmdtip2'}
See how dairy has the highest alpha? This means it produces the most $P$ of all 4 classes.
```

#### Calculate phosphate export load {#M5B_8 -}

6) Compute the $P_{load}$ at the catchment outlet in kg/day. Also, work out the cumulative load (this will come in handy to coampre scenarios; we call it a mass curve) 

#### Consider 2 management scenarios {#M5B_9 -}

Finally, let’s run 2 management scenarios:

7a) **Scenario 1:** If the total urban area expanded to 50% (0.5), how would this affect the overall $P$ export?

```{block2, hintM5B_4, type='rmdtip2'}
Change the land use fractions manually. They still have to add up to 1, so change the other parts as you see appropriate, i.e. urban area usually comes at the expense of forest, but you can decide how to change the landscape. Record your changed fractions.
```

7b) **Scenario 2** If the irrigated dairy industry were made to contain their flows (i.e. prevent high $P$ concentrations going to the stream), how would this affect the overall concentration in the water at the river basin outlet? The total concentration has to be maintained at 0.065 g m^3^ or below. 

```{block2, hintM5B_5, type='rmdtip2'}
Use an IF statement.
```

### Submission {#M5B_10 -}
:::: {.redbox2}

**Submit properly formatted graphs and tables of the following sections of the lab:**

1) Clear photograph or page scan of your hand drawn conceptual model of the Catchment system (may be combined with the 5a catchment model)
2) Plot of calibration regression plot, including a trendline, equation and R^2^ value. Include in the caption what parameter values you gave the catchment.
3) Plot of the cumulative phosphorous export of increased urbanisation (Scenario 1) and dairy management (Scenario 2) compared to the original conditions, over time.
4) Plot of the concentration-flow relationship for the three scenarios of phosphorous concentration. 
5) Any other interesting plots you may like to create.

These are in a PDF format, combined with 5a results. No screenshots of figures from Excel/Excel spreadsheets to be uploaded

**General professional formatting guidelines:**

- All figures are to have adequate captions explaining them
- For graphs, figure captions go below the plot
- For tables, the caption goes above the table
- Make sure figures and their text size is readable

**Excel hints:**

- When there is a caption for a plot, you remove the title
- Remove the plot border and gridlines
- Make sure both axes have visible lines and tick marks
- Units need to be noted properly with the axis label - 'Temperature (°C)'
- Round numbers to be reasonable
::::

